\\textsc{Mathematical Description}

### 2022-03-23 100-math-description/GR-sslcm-math-description.Rmd

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: false
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
fontfamily: mathpazo
geometry: margin=1in
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
---

@@

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \rightarrow B}_{y,j}$ as an example.
We denote its random variability across years and populations as:
\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \rightarrow B}_{y,1:n_j}\right), \Sigma_{\phi^{A \rightarrow B}} \right]
(\#eq:generic-random)
\end{equation}
where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix with the structure:
\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma4by4)
\end{equation}
That is, all populations ($j$) have unique variance terms $(\sigma_{\phi^{A \rightarrow B}_j})^2$ but we assume the correlation component of the covariance calculation ($\rho_{\phi^{A \rightarrow B}}$) is identical for all population pairs.
In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:
\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \rightarrow B}_{o=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{o=2}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma2by2)
\end{equation}
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \@ref(tab:param-table)) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \@ref(eq:generic-Sigma4by4) and \@ref(eq:generic-Sigma2by2).

@@

Total egg production is the sum product of age-specific spawner abundance (from eq. \@ref(eq:get-Sa), below), proportion female, and fecundity:
\begin{equation}
E_{y,j} = \sum_o^{n_o} \sum_k^{n_k} S^{a}_{y,k,o,j} \cdot \Omega_{k,j} \cdot f_{k}
(\#eq:get-E)
\end{equation}
The proportion of spawners at age that are female ($\Omega_{k,j}$) was calculated from carcass survey data and averaged across years and fecundity ($f_k$) was calculated from samples taken from Grande Ronde-origin hatchery broodstock.
Both quantities were supplied to the model assuming no error or inter-annual variability.

Deterministic egg-to-parr survival is density-dependent and follows a Beverton-Holt relationship with productivity parameter $\alpha_j$ and capacity parameter $\beta_j$:
\begin{equation}
\dot{\phi}^{E \rightarrow P^{b}}_{y,j} = \frac{1}{\frac{1}{\alpha_j} + \frac{E_{y,j}}{\beta_j}}
(\#eq:get-phi-E-Pb-dot)
\end{equation}
To facilitate later analyses (not documented here) investigating the effects of habitat restoration and climate change on population dynamics, we modeled parr capacity as a function of weighted usable habitat length ($\mathrm{WUL}_j$).
Derivation of $\mathrm{WUL}_j$ is too detailed to describe fully here, but briefly, the "value" of stream segments with different habitat characteristics were quantified by a complex generalized linear mixed model that treated detectability-adjusted snorkel counts as the response variable and habitat features (e.g., pool frequency, water temperature, large wood density, etc.) as predictor variables and weights were assigned to stream segments based on their relative value to the segment with the highest value; $\mathrm{WUL}_j$ was then the sum of segment-specific weight and length within the known spawning and rearing extent for each population. We modeled capacity ($\beta_j$) as a function of this $\mathrm{WUL}_j$ metric:
\begin{equation}
\log(\beta_j) \sim \mathcal{N}\left[\log\left(\lambda \cdot \mathrm{WUL}_j\right), \sigma_{\beta} \right]
(\#eq:beta-random)
\end{equation}
where $\lambda$ is the expected change in parr capacity per 1 km change in weighted usable habitat length, and $\sigma_{\beta}$ is the log-normal standard deviation of variability in this relationship not captured by $\mathrm{WUL}$ values.

Realized (i.e., with process noise) egg-to-parr survival was a multivariate logit-normal random variable around the deterministic value ($\dot{\phi}^{E \rightarrow P^{b}}_{y,j}$) predicted by the Beverton-Holt function in eq. \@ref(eq:get-phi-E-Pb-dot) with covariance matrix $\Sigma_{\phi^{E \rightarrow P^{b}}}$ (constructed of $\sigma_{\phi^{E \rightarrow P^{b}}_j}$ and $\rho_{\phi^{E \rightarrow P^{b}}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{E \rightarrow P^{b}}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{E \rightarrow P^{b}}_{y,1:n_j}\right), \Sigma_{\phi^{E \rightarrow P^{b}}}\right]
(\#eq:phi-E-Pb-random)
\end{equation}
Parr recruitment ($P^{b}_{y,j}$) to the end of summer was then the product of total egg production and egg-to-parr survival:
\begin{equation}
P^{b}_{y,j} = \phi^{E \rightarrow P^{b}}_{y,j} \cdot E_{y,j}
(\#eq:get-Pb)
\end{equation}

@@

Not all summer parr ($P^{b}_{y,j}$) stay in the headwaters to rear overwinter -- some migrate out in the fall and rear farther downstream (termed "fall migrants" and indexed by $i = \mathrm{fall}$) and others migrate out the following spring (termed "spring migrants" and indexed by $i = \mathrm{spring}$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately.
To enable building an observation model that could accommodate this feature of the populations in our model, summer parr was apportioned to these two strategies:
\begin{equation}
P^{a}_{y,i,j} = P^{b}_{y,j} \cdot \pi_{y,i,j}
(\#eq:get-Pa)
\end{equation}
where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
For fall migrants, this proportion was a multivariate logit-normal random variable around a time-constant mean ($\dot{\pi}_{i = \mathrm{fall},j}$) with covariance matrix $\Sigma_{\pi_{i = \mathrm{fall}}}$ (constructed of $\sigma_{\pi_{i=\mathrm{fall},j}}$ and $\rho_{\pi_{i=\mathrm{fall}}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\pi}_{y,i = \mathrm{fall},1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\pi}}_{i = \mathrm{fall},1:n_j}\right), \Sigma_{\pi_{i = \mathrm{fall}}}\right]
(\#eq:pi-random)
\end{equation}
and the proportion that are spring migrants was obtained as its complement ($\pi_{y,i = \mathrm{spring},j} = 1 - \pi_{y,i = \mathrm{fall},j}$).

@@

Overwinter survival from the end of the parr year (i.e., age-1) to the start of the smolt year (i.e., age-2), denoted by $\phi^{P^{a} \rightarrow M^{b}}_{y,i,j}$, was modeled as a logit-linear function of parr density (scaled to the same metric of weighted usable habitat length [$\mathrm{WUL}_j$] used in eq. \@ref(eq:beta-random)).
Deterministic overwinter survival was:
\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,j} \cdot \left(\frac{P^{a}_{y,i,j}}{\mathrm{WUL}_j}\right)
(\#eq:get-phi-Pa-Mb-dot)
\end{equation}
Note the migratory strategy-specific intercept ($\gamma_{0,i,j}$) but common slope terms ($\gamma_{1,j}$).
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the deterministic logistic model prediction with covariance matrix $\Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}$ (each matrix constructed of $\sigma_{\phi^{P^{a} \rightarrow M^{b}}_{i,j}}$ and $\rho_{\phi^{P^{a} \rightarrow M^{b}}_{i}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}\right]
(\#eq:phi-Pa-Mb-random)
\end{equation}
Natural-origin (indexed by $o=\mathrm{NOR}$) parr surviving to the beginning of the following year was then:
\begin{equation}
M^{b}_{y,i,o=\mathrm{NOR},j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \rightarrow M^{b}}_{y,i,j}
(\#eq:get-Mb-NOR)
\end{equation}
For populations with hatchery supplementation, hatchery-origin smolt releases (assumed known without error) were introduced at this point as well and stored in the variable $M^{b}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}$; we assumed no hatchery-origin parr were released the prior year, thus $M^{b}_{y,i=\mathrm{fall},o=\mathrm{HOR},j} = 0$.

@@

In-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2.
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{fall},o=\mathrm{NOR},j} = \phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}$, we thus omit the $i$ dimension when defining these parameter below), but we were able to model natural-origin survival separately from hatchery-origin.
We assume that year-specific realized survival values were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{b} \rightarrow M^{a}}_{o,j}$) with covariance matrix $\Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}$ (constructed of $\sigma_{\phi^{M^{b} \rightarrow M^{a}}_{o,j}}$ and $\rho_{\phi^{M^{b} \rightarrow M^{a}}_{o}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \rightarrow M^{a}}_{o,1:n_j}\right), \Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}\right]
(\#eq:phi-Mb-Ma-random)
\end{equation}
Thus, the abundance of smolt reaching LGR was:
\begin{equation}
M^{a}_{y,i,o,j} = M^{b}_{y,i,o,j} \cdot \phi^{M^{b} \rightarrow M^{a}}_{y,o,j}
(\#eq:get-Ma)
\end{equation}
After reaching LGR, we modeled survival during the downstream migration through the hydrosystem on the Snake and Columbia rivers as a single survival rate, specific to each year and origin but shared among populations ($\phi^{M^{a} \rightarrow O^{0}}_{y,o}$).
We assumed these survival rates were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{a} \rightarrow O^{0}}_{o}$) with covariance matrix $\Sigma_{\phi^{M^{a} \rightarrow O^{0}}}$ (constructed of $\sigma_{\phi^{M^{a} \rightarrow O^{0}}_{o}}$ and $\rho_{\phi^{M^{a} \rightarrow O^{0}}}$ following eq. \@ref(eq:generic-Sigma2by2)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{a} \rightarrow O^{0}}_{y,1:n_o}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{a} \rightarrow O^{0}}_{1:n_o}\right), \Sigma_{\phi^{M^{a} \rightarrow O^{0}}}\right]
(\#eq:phi-Ma-O0-random)
\end{equation}
Thus, the abundance of fish reaching the ocean by origin was:
\begin{equation}
O^{0}_{y,o,j} = \phi^{M^{a} \rightarrow O^{0}}_{y,o} \sum_i^{n_i} M^{a}_{y,i,o,j}
(\#eq:get-O0)
\end{equation}
Note the summation across migratory strategies ($i$) in eq. \@ref(eq:get-O0), which indicates that the fates of fish with these different strategies were assumed to be the result of identical processes following arrival to LGR.

@@

\begin{equation}
\boldsymbol{\hat{x}}^{R^{a}}_{y,1:n_{ko},j} \sim \mathcal{M}\left(\boldsymbol{p}^{R^{a}}_{y,1:n_{ko},j}, \sum_{ko}^{n_{ko}} \hat{x}^{R^{a}}_{y,ko,j}\right)
(\#eq:weir-comp-data)
\end{equation}
Due to the age sampling bias of carcass surveys we discovered, we required a correction to ensure reliable fits to all data sets while still recovering unbiased true age composition data for the Minam River population, which had only carcass composition data (the other three populations had both weir and carcass composition data).

>>>>

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: false
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: margin=1in
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
---

@@

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \rightarrow B}_{y,j}$ as an example.
We denote its random variability across years and populations as:

\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \rightarrow B}_{y,1:n_j}\right), \Sigma_{\phi^{A \rightarrow B}} \right]
(\#eq:generic-random)
\end{equation}

where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix with the structure:

\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma4by4)
\end{equation}

That is, all populations ($j$) have unique variance terms $(\sigma_{\phi^{A \rightarrow B}_j})^2$ but we assume the correlation component of the covariance calculation ($\rho_{\phi^{A \rightarrow B}}$) is identical for all population pairs.
In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:

\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \rightarrow B}_{o=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{o=2}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma2by2)
\end{equation}

Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \@ref(tab:param-table)) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \@ref(eq:generic-Sigma4by4) and \@ref(eq:generic-Sigma2by2).

@@

Total egg production is the sum product of age-specific spawner abundance (from eq. \@ref(eq:get-Sa), below), proportion female, and fecundity:

\begin{equation}
E_{y,j} = \sum_o^{n_o} \sum_k^{n_k} S^{a}_{y,k,o,j} \cdot \Omega_{k,j} \cdot f_{k}
(\#eq:get-E)
\end{equation}

The proportion of spawners at age that are female ($\Omega_{k,j}$) was calculated from carcass survey data and averaged across years and fecundity ($f_k$) was calculated from samples taken from Grande Ronde-origin hatchery broodstock.
Both quantities were supplied to the model assuming no error or inter-annual variability.

Deterministic egg-to-parr survival is density-dependent and follows a Beverton-Holt relationship with productivity parameter $\alpha_j$ and capacity parameter $\beta_j$:

\begin{equation}
\dot{\phi}^{E \rightarrow P^{b}}_{y,j} = \frac{1}{\frac{1}{\alpha_j} + \frac{E_{y,j}}{\beta_j}}
(\#eq:get-phi-E-Pb-dot)
\end{equation}

To facilitate later analyses (not documented here) investigating the effects of habitat restoration and climate change on population dynamics, we modeled parr capacity as a function of weighted usable habitat length ($\mathrm{WUL}_j$).
Derivation of $\mathrm{WUL}_j$ is too detailed to describe fully here, but briefly, the "value" of stream segments with different habitat characteristics were quantified by a complex generalized linear mixed model that treated detectability-adjusted snorkel counts as the response variable and habitat features (e.g., pool frequency, water temperature, large wood density, etc.) as predictor variables and weights were assigned to stream segments based on their relative value to the segment with the highest value; $\mathrm{WUL}_j$ was then the sum of segment-specific weight and length within the known spawning and rearing extent for each population. We modeled capacity ($\beta_j$) as a function of this $\mathrm{WUL}_j$ metric:

\begin{equation}
\log(\beta_j) \sim \mathcal{N}\left[\log\left(\lambda \cdot \mathrm{WUL}_j\right), \sigma_{\beta} \right]
(\#eq:beta-random)
\end{equation}

where $\lambda$ is the expected change in parr capacity per 1 km change in weighted usable habitat length, and $\sigma_{\beta}$ is the log-normal standard deviation of variability in this relationship not captured by $\mathrm{WUL}$ values.

Realized (i.e., with process noise) egg-to-parr survival was a multivariate logit-normal random variable around the deterministic value ($\dot{\phi}^{E \rightarrow P^{b}}_{y,j}$) predicted by the Beverton-Holt function in eq. \@ref(eq:get-phi-E-Pb-dot) with covariance matrix $\Sigma_{\phi^{E \rightarrow P^{b}}}$ (constructed of $\sigma_{\phi^{E \rightarrow P^{b}}_j}$ and $\rho_{\phi^{E \rightarrow P^{b}}}$ following eq. \@ref(eq:generic-Sigma4by4)):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{E \rightarrow P^{b}}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{E \rightarrow P^{b}}_{y,1:n_j}\right), \Sigma_{\phi^{E \rightarrow P^{b}}}\right]
(\#eq:phi-E-Pb-random)
\end{equation}

Parr recruitment ($P^{b}_{y,j}$) to the end of summer was then the product of total egg production and egg-to-parr survival:

\begin{equation}
P^{b}_{y,j} = \phi^{E \rightarrow P^{b}}_{y,j} \cdot E_{y,j}
(\#eq:get-Pb)
\end{equation}

@@

Not all summer parr ($P^{b}_{y,j}$) stay in the headwaters to rear overwinter -- some migrate out in the fall and rear farther downstream (termed "fall migrants" and indexed by $i = \mathrm{fall}$) and others migrate out the following spring (termed "spring migrants" and indexed by $i = \mathrm{spring}$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately.
To enable building an observation model that could accommodate this feature of the populations in our model, summer parr was apportioned to these two strategies:

\begin{equation}
P^{a}_{y,i,j} = P^{b}_{y,j} \cdot \pi_{y,i,j}
(\#eq:get-Pa)
\end{equation}

where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
For fall migrants, this proportion was a multivariate logit-normal random variable around a time-constant mean ($\dot{\pi}_{i = \mathrm{fall},j}$) with covariance matrix $\Sigma_{\pi_{i = \mathrm{fall}}}$ (constructed of $\sigma_{\pi_{i=\mathrm{fall},j}}$ and $\rho_{\pi_{i=\mathrm{fall}}}$ following eq. \@ref(eq:generic-Sigma4by4)):

\begin{equation}
\logit\left(\boldsymbol{\pi}_{y,i = \mathrm{fall},1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\pi}}_{i = \mathrm{fall},1:n_j}\right), \Sigma_{\pi_{i = \mathrm{fall}}}\right]
(\#eq:pi-random)
\end{equation}

and the proportion that are spring migrants was obtained as its complement ($\pi_{y,i = \mathrm{spring},j} = 1 - \pi_{y,i = \mathrm{fall},j}$).

@@

Overwinter survival from the end of the parr year (i.e., age-1) to the start of the smolt year (i.e., age-2), denoted by $\phi^{P^{a} \rightarrow M^{b}}_{y,i,j}$, was modeled as a logit-linear function of parr density (scaled to the same metric of weighted usable habitat length [$\mathrm{WUL}_j$] used in eq. \@ref(eq:beta-random)).
Deterministic overwinter survival was:

\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,j} \cdot \left(\frac{P^{a}_{y,i,j}}{\mathrm{WUL}_j}\right)
(\#eq:get-phi-Pa-Mb-dot)
\end{equation}

Note the migratory strategy-specific intercept ($\gamma_{0,i,j}$) but common slope terms ($\gamma_{1,j}$).
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the deterministic logistic model prediction with covariance matrix $\Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}$ (each matrix constructed of $\sigma_{\phi^{P^{a} \rightarrow M^{b}}_{i,j}}$ and $\rho_{\phi^{P^{a} \rightarrow M^{b}}_{i}}$ following eq. \@ref(eq:generic-Sigma4by4)):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}\right]
(\#eq:phi-Pa-Mb-random)
\end{equation}

Natural-origin (indexed by $o=\mathrm{NOR}$) parr surviving to the beginning of the following year was then:

\begin{equation}
M^{b}_{y,i,o=\mathrm{NOR},j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \rightarrow M^{b}}_{y,i,j}
(\#eq:get-Mb-NOR)
\end{equation}

For populations with hatchery supplementation, hatchery-origin smolt releases (assumed known without error) were introduced at this point as well and stored in the variable $M^{b}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}$; we assumed no hatchery-origin parr were released the prior year, thus $M^{b}_{y,i=\mathrm{fall},o=\mathrm{HOR},j} = 0$.

@@

In-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2.
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{fall},o=\mathrm{NOR},j} = \phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}$, we thus omit the $i$ dimension when defining these parameter below), but we were able to model natural-origin survival separately from hatchery-origin.
We assume that year-specific realized survival values were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{b} \rightarrow M^{a}}_{o,j}$) with covariance matrix $\Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}$ (constructed of $\sigma_{\phi^{M^{b} \rightarrow M^{a}}_{o,j}}$ and $\rho_{\phi^{M^{b} \rightarrow M^{a}}_{o}}$ following eq. \@ref(eq:generic-Sigma4by4)):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \rightarrow M^{a}}_{o,1:n_j}\right), \Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}\right]
(\#eq:phi-Mb-Ma-random)
\end{equation}

Thus, the abundance of smolt reaching LGR was:

\begin{equation}
M^{a}_{y,i,o,j} = M^{b}_{y,i,o,j} \cdot \phi^{M^{b} \rightarrow M^{a}}_{y,o,j}
(\#eq:get-Ma)
\end{equation}

After reaching LGR, we modeled survival during the downstream migration through the hydrosystem on the Snake and Columbia rivers as a single survival rate, specific to each year and origin but shared among populations ($\phi^{M^{a} \rightarrow O^{0}}_{y,o}$).
We assumed these survival rates were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{a} \rightarrow O^{0}}_{o}$) with covariance matrix $\Sigma_{\phi^{M^{a} \rightarrow O^{0}}}$ (constructed of $\sigma_{\phi^{M^{a} \rightarrow O^{0}}_{o}}$ and $\rho_{\phi^{M^{a} \rightarrow O^{0}}}$ following eq. \@ref(eq:generic-Sigma2by2)):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{a} \rightarrow O^{0}}_{y,1:n_o}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{a} \rightarrow O^{0}}_{1:n_o}\right), \Sigma_{\phi^{M^{a} \rightarrow O^{0}}}\right]
(\#eq:phi-Ma-O0-random)
\end{equation}

Thus, the abundance of fish reaching the ocean by origin was:

\begin{equation}
O^{0}_{y,o,j} = \phi^{M^{a} \rightarrow O^{0}}_{y,o} \sum_i^{n_i} M^{a}_{y,i,o,j}
(\#eq:get-O0)
\end{equation}

Note the summation across migratory strategies ($i$) in eq. \@ref(eq:get-O0), which indicates that the fates of fish with these different strategies were assumed to be the result of identical processes following arrival to LGR.

@@

\begin{equation}
\boldsymbol{\hat{x}}^{R^{a}}_{y,1:n_{ko},j} \sim \mathcal{M}\left(\boldsymbol{p}^{R^{a}}_{y,1:n_{ko},j}, \sum_{ko}^{n_{ko}} \hat{x}^{R^{a}}_{y,ko,j}\right)
(\#eq:weir-comp-data)
\end{equation}

Due to the age sampling bias of carcass surveys we discovered, we required a correction to ensure reliable fits to all data sets while still recovering unbiased true age composition data for the Minam River population, which had only carcass composition data (the other three populations had both weir and carcass composition data).
For the three populations with paired composition data (i.e., $j \in [\mathrm{CAT}, \mathrm{LOS}, \mathrm{UGR}]$), we estimated correction factors in a hierarchical fashion:

\begin{equation}
z_{k,j} \sim \mathcal{N}(\dot{z}_{k}, \sigma_{z_k})
(\#eq:z-random)
\end{equation}
